---
title: "H3K4 methylation at broad promoters and enhancers"
author: "Aleksandra Pekowska"
output:
  html_document:
    highlight: tango
    theme: flatly
    toc: yes
    toc_depth: 3
  pdf_document:
    toc: yes
    toc_depth: '3'
---

<style type="text/css">

h1.title {
  font-size: 30px;
  color: DarkBlue;
  text-align: center;
}
h4.author { /* Header 4 - and the author and data headers use this too  */
    font-size: 20px;
  text-align: center;
}
</style>


# Preparations

Load libraries, set the paths.
The HMD corrected files were taken from **https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE103543**. The H3K4me3 file was taken from: **https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM2773406**. 

```{r, echo=FALSE, results='hide', warning=FALSE}

library(rtracklayer)
library(biomaRt)
library(ggplot2)
library(data.table)
library(Gviz)
library(stringr)
library(EnsDb.Hsapiens.v86)
library(topGO)
library(VennDiagram)
library(ggVennDiagram)
library(GO.db)
library(ensembldb)
set.seed(22)
edb = EnsDb.Hsapiens.v86
goterms = Term(GOTERM)

peak_dir = "/Volumes/Expansion/Pekowska_et_al_MolCell2021/peaks/"
wig_dir = "/Volumes/Expansion/Pekowska_et_al_MolCell2021/bigWigs/"
objects = "/Volumes/Expansion/Pekowska_et_al_MolCell2021/objects/"
figures = "/Volumes/Expansion/Pekowska_et_al_MolCell2021/figures/"
source = "/Volumes/Expansion/Pekowska_et_al_MolCell2021//source/"

```


# Gene annotation

Throughout the methods, we consider one annotation of genes (*ensembl*).


```{r,eval=TRUE}
ensembl = useEnsembl(biomart = 'genes', 
                     dataset = 'hsapiens_gene_ensembl',
                     version = 95)
searchDatasets(mart = ensembl, pattern = "hsapiens_gene_ensembl")

filt=c(1:22,'X','Y')

loci = getBM(attributes= c("refseq_mrna",
                           "external_gene_name",
                           "chromosome_name",
                           "strand",
                           "transcription_start_site",
                           "transcript_start",
                           "transcript_end",
                           "start_position",
                           "end_position",
                           "gene_biotype"),
             filters=c("chromosome_name"),
             values=filt, 
             mart=ensembl)
length(unique(loci$refseq_mrna))
loci$STRAND = rep('-',nrow(loci))
loci$STRAND[loci$strand==1] = '+'
loci$chr = paste0('chr',loci$chromosome_name)

table( loci$gene_biotype )
all( loci$transcript_start < loci$transcript_end )
```


Now generate TSS and gene ranges. 


```{r, warning=FALSE}

tss = GRanges( seqnames = Rle(loci$chr),
               ranges = IRanges(as.numeric(loci$transcription_start_site )-2000,
                                end=as.numeric(loci$transcription_start_site )+2000),
               strand = rep('*', nrow(loci)) )

genes = GRanges( seqnames = Rle(loci$chr),
                 ranges = IRanges(as.numeric(loci$start_position )-2000,
                                  end=as.numeric(loci$end_position )+2000),
                 strand = rep('*', nrow(loci)) )

```


# Analysis of enhancers 

Note - all the files were mapped to the GRCh38 version of the human genome.
We will consider only the High confidence peaks provided by the authors. Below we indicate which files were used in the analyses.

```{r, warning=FALSE}

readBed_filterChroms = function(filePath, chroms, scoreCol){
  ## filePath="/Volumes/ALEKSANDRA/carte_SD/Re_Shah_et_al/peaks/ENCFF516BJP_optimal_idr_thresholded_peaks_reps_1_and_2.bed"; chroms=paste0('chr',c(1:22,'X','Y'));scoreCol=10
  x= read.delim(filePath,as.is=T,header=F)
  x = x[x[,1] %in% chroms,]
return( GRanges( seqnames = Rle(x[,1]),
               ranges = IRanges(as.numeric(x[,2]),
                                end=as.numeric(x[,3]),
                                names=rownames(x)),
               score=x[,scoreCol]) ) } 

me1_peaks = readBed_filterChroms( paste0(peak_dir, "GSE103543_AR16_H3K4me1-Corrected_High_Confidence_Peaks.bed"), 
                                  chroms=paste0('chr',c(1:22,'X','Y')), 4 )

me3_peaks = readBed_filterChroms(paste0(peak_dir,
                              "GSM2773406_AR16-4_TF-PA5-40086_High_Confidence_Peaks.bed"),
                       chroms=paste0('chr',c(1:22,'X','Y')), 4 )


pol=readBed_filterChroms(paste0(peak_dir,
                       "ENCFF516BJP_optimal_idr_thresholded_peaks_reps_1_and_2.bed"),
                       chroms=paste0('chr',c(1:22,'X','Y')), 4) 
               

```

## Identifincation of enhancers

We considered two most recent DHS experiments from John Stamatoyannopoulos lab and worked with DHS peaks that were detected in the two replicate experiments. 

```{r, warning=FALSE}

dhs_rep1 = readBed_filterChroms( paste0(peak_dir, "ENCFF100IJK_DHS_Stam_2018_rep1.bed"),
                                 chroms=paste0('chr',c(1:22,'X','Y')), 4)
seqlevelsStyle(dhs_rep1) = 'ncbi'


dhs_rep2 = readBed_filterChroms( paste0(peak_dir, "ENCFF339PKG_DHS_Stam_2018_rep2.bed"),
                                 chroms=paste0('chr',c(1:22,'X','Y')), 4)
seqlevelsStyle(dhs_rep2) = 'ncbi'

dhs = dhs_rep1[ queryHits(findOverlaps(dhs_rep1,dhs_rep2) ) ]
dhs = reduce(dhs)

seqlevelsStyle(dhs) = 'ucsc'

dhs
```


We have **54726** high confidence DHS sites in K562. 

## Identification of **distal enhancers**


We will consider only the DHS tht overlap H3K4me1 peaks.

```{r}
enhancers = dhs[ unique(queryHits( findOverlaps(dhs, me1_peaks)) ) ]
```


We have **38588** putative enhancers. We will next identify **DHS that are distal with respect to the TSS** in the the human genome. Note that we will **consider all the RNA species annotated so far, including lincRNAs and others**.


```{r, warning=FALSE}
enhancers = enhancers[ -unique(queryHits( findOverlaps(enhancers, tss)) ) ]
```


Based on the TSS annotation, we identify **24649 DHS** that we will consider in the further steps of the analysis. We **next filter these instances for elements that do not intersect genes**.


```{r, warning=FALSE}
distal_enhancers = enhancers[ -unique(queryHits( findOverlaps(enhancers, genes)) )  ]
```


Now we have 9155 elements to consider. Export bed to execute the "lift-over" to *Hg19* using ucsc.

```{r}
distal_enhancers
```

```{r, warning=FALSE}

export.bed(distal_enhancers,paste0(objects,"distal_enhancers_GRCh38.bed"))

distal_enhancers_hg19 = readBed_filterChroms( paste0(objects,"distal_enhancers_GRCh37.bed"),
                                 chroms=paste0('chr',c(1:22,'X','Y')), 5)

```


### Active and not enhancers

Identify enhancers with and without pol2 peaks.


```{r, warning=FALSE}

distal_enhancers_no_pol = distal_enhancers[ -unique(queryHits(findOverlaps(distal_enhancers,pol)))]
distal_enhancers_with_pol = distal_enhancers[ unique(queryHits(findOverlaps(distal_enhancers,pol)))]

save( distal_enhancers_no_pol, distal_enhancers_with_pol, 
      file=paste0(objects, 'Enhancers_Pol2_stratification_GRCh38.RData'))

```


Statistics.


```{r, warning=FALSE, Distal_enhancers_pol2, fig.width=10, fig.height=10, out.width='50%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

statsP = c( length(unique(queryHits(findOverlaps(distal_enhancers_no_pol,me3_peaks)))),
           length(unique(queryHits(findOverlaps(distal_enhancers_with_pol,me3_peaks) ))) )

statsP

statsP = statsP/c(length(distal_enhancers_no_pol),
                  length(distal_enhancers_with_pol))
statsP = statsP*100

barplot( statsP, col=c('red','blue'),
         names=c('Without','With'), 
         ylim=c(0,25),
         ylab='% of distal enhancers intersecting H3K4me3 peak', 
         xlab='Pol2 peak' )

```


### Active and inactive enhancers - GRO-seq based classification

Next, we score transcription at these distal DHS sites.


```{r, warning=FALSE}

gro_min = import.bw( paste0(wig_dir,"GSM1480325_K562_GROseq_minus.bigWig") )
gro_plus= import.bw( paste0(wig_dir,"GSM1480325_K562_GROseq_plus.bigWig") )

gro = c(gro_min,gro_plus)

##Add the minus strand too
gro$score = as.numeric( abs(gro$score) )

```


Now score for expression level at the distal enhancers.


```{r, warning=FALSE, distal_enhancers_GROseq_signal, fig.width=15, fig.height=10, out.width='80%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

distal_enhancers_no_expression = distal_enhancers_hg19[ -unique(queryHits(findOverlaps(distal_enhancers_hg19,gro)))]

gro_score_distal_enhancers = as.data.frame( findOverlaps(distal_enhancers_hg19,gro) )
gro_score_distal_enhancers$gro = score(gro)[ gro_score_distal_enhancers[,2] ]
gro_score_distal_enhancers = gro_score_distal_enhancers[gro_score_distal_enhancers$gro>0,]

gro_score_distal_enhancers = split(gro_score_distal_enhancers[,3], gro_score_distal_enhancers[,1])

distal_enhancers_hg19_transcription = unlist(lapply(gro_score_distal_enhancers,function(x){sum(as.numeric(x),na.rm=TRUE)}))


hist( log2(distal_enhancers_hg19_transcription),n=20, 
      main='Enhancers with GRO-seq signal', col='gray60',
      xlab=expression("GRO-seq number of reads (log"[2]*")"))
abline(v=2,col='red',lwd=2)


distal_enhancers_low_expression = distal_enhancers_hg19[ as.numeric(names(distal_enhancers_hg19_transcription[distal_enhancers_hg19_transcription<=2]))]

distal_enhancers_high_expression = distal_enhancers_hg19[ as.numeric(names(distal_enhancers_hg19_transcription[distal_enhancers_hg19_transcription>2]))]


```


Export the files and lift them over to GRCh38.


```{r, warning=FALSE}

export.bed(distal_enhancers_no_expression,
           paste0(objects, "distal_enhancers_no_expression_GRCh37.bed") )
export.bed(distal_enhancers_low_expression,
           paste0(objects, "distal_enhancers_low_expression_GRCh37.bed") )
export.bed(distal_enhancers_high_expression,
           paste0(objects, "distal_enhancers_high_expression_GRCh37.bed") )


readBed_filterChroms( paste0(objects,"distal_enhancers_GRCh37.bed"),
                                 chroms=paste0('chr',c(1:22,'X','Y')), 5)

distal_enhancers_no_expression_38=readBed_filterChroms(paste0(objects,"distal_enhancers_no_expression_GRCh38.bed"), chroms=paste0('chr',c(1:22,'X','Y')), 5)
distal_enhancers_low_expression_38=readBed_filterChroms(paste0(objects,"distal_enhancers_low_expression_GRCh38.bed"), chroms=paste0('chr',c(1:22,'X','Y')), 5)
distal_enhancers_high_expression_38=readBed_filterChroms(paste0(objects,"distal_enhancers_high_expression_GRCh38.bed"), chroms=paste0('chr',c(1:22,'X','Y')), 5)

save( distal_enhancers_high_expression_38,
      distal_enhancers_low_expression_38,
      distal_enhancers_no_expression_38,
      file= paste0(objects, 'Enhancers_GRO_stratification_GRCh38.RData' ) )

```


### Enhancers overlap with H3K4me3

How many of the distal DHS sites (stratified by expression measured by GRO-seq), overlap H3K4me3 peaks.


```{r, warning=FALSE, distal_enhancers_overlap_with_H3K4, fig.width=10, fig.height=8, out.width='50%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

load( paste0(objects, 'Enhancers_GRO_stratification_GRCh38.RData' ) )
stats = c( length(unique(queryHits(findOverlaps(distal_enhancers_no_expression_38,me3_peaks)))),
           length(unique(queryHits(findOverlaps(distal_enhancers_low_expression_38,me3_peaks) ))),
           length( unique(queryHits(findOverlaps(distal_enhancers_high_expression_38,me3_peaks) )) ))

stats

stats = stats/c(length(distal_enhancers_no_expression_38),
                length(distal_enhancers_low_expression_38),
                length(distal_enhancers_high_expression_38))
stats = stats*100

barplot( stats, col=c('red','blue','blue4'),
         names=c('No','Low','High'), ylim=c(0,20),
         ylab='% of distal enhancers intersecting H3K4me3 peak', 
         xlab='GRO-seq expression level' )

```


### Visualisation 

Let's display some examples in the browser. We choose the most specific antibody (H3K4me3-3), the high confidence H3K4me3 peaks and the annotation of distal enhancers.


```{r}

me3_track_HMD = import( paste0(wig_dir,'GSM2773406_AR16-4_TF-PA5-40086_mQ20_L200_HMD.bigwig' ) )
seqlevelsStyle(me3_track_HMD)='ncbi'

```


Plot examples.

The GATA1 locus.


```{r, plotGATA1, fig.width=12, fig.height=10, out.width='50%', fig.align='center', dev.args=list(pointsize=30),fig.path=figures, dev=c('png', 'pdf'), warning=FALSE}
options(ucscChromosomeNames=FALSE)
st = 48775000+7000
en = 48787500
thechr='X'

bedgraph_dt_one_chr1 = me3_track_HMD[ which( seqnames(me3_track_HMD) == thechr ) ]
me3_peaks_ncbi=me3_peaks
seqlevelsStyle(me3_peaks_ncbi)='ncbi'

distal_enhancers_ncbi=distal_enhancers
seqlevelsStyle(distal_enhancers_ncbi)='ncbi'

biomTrack = BiomartGeneRegionTrack(genome="hg38", biomart=ensembl,
                                   chromosome=thechr, 
                                   start=st, end=en,
                                   name="ENSEMBL",
                                   background.panel = "#FFFFFF",
                                   background.title = "#FFFFFF",
                                   col.axis="black",
                                   col.title="black",col=1,
                                   fill='gray70',
                                   ontfamily="Arial",
                                   showID = FALSE, geneSymbol = TRUE)


## ChIPseq
dtrac1 = DataTrack(range = bedgraph_dt_one_chr1,
                   type = "h",
                   genome = 'hg38',
                   name = "H3K4me3 HMD (40086)",
                   background.panel = "#FFFFFF",
                   background.title = "#FFFFFF",
                   col.axis="black",
                   col.title="black",col=1,
                   fill=1,
                   fontfamily="Arial" )

## peaks
thisReg_me3_peaks1 = AnnotationTrack(me3_peaks_ncbi, genome='hg38',
                                     name='H3K4me3',
                                     chromosome = thechr,shape='box',
                                     fill='blue',size=1.5,
                                     background.panel = "#FFFFFF",
                                     background.title = "#FFFFFF",
                                     col.axis="black",
                                     col.title="black",col=1,
                                     fontfamily="Arial")

thisReg_enhancers = AnnotationTrack(distal_enhancers_ncbi, genome='hg38',
                                    name='Distal Enhancers',
                                    chromosome = thechr,shape='box',
                                    fill='red',size=1.5, 
                                    background.panel = "#FFFFFF",
                                    background.title = "#FFFFFF",
                                    col.axis="black",
                                    col.title="black",col=1,
                                    fontfamily="Arial")


# itrack = IdeogramTrack(genome = "hg38", chromosome = thechr)
gtrack = GenomeAxisTrack()


plotTracks(list(dtrac1,
                thisReg_me3_peaks1, 
                thisReg_enhancers,
                biomTrack),
           chromosome = thechr, 
           from = st, to = en)

```



The profiles along with annotations at the PRKAR2B locus.

```{r, plotPRKAR2B, fig.width=12, fig.height=10, out.width='50%', fig.align='center', dev.args=list(pointsize=30),fig.path=figures, dev=c('png', 'pdf'), warning=FALSE}
thechr = 7
st = 106960000+35000
en = 107060000
bedgraph_dt_one_chr1 = me3_track_HMD[ which( seqnames(me3_track_HMD) == thechr ) ]
me3_peaks_ncbi=me3_peaks
seqlevelsStyle(me3_peaks_ncbi)='ncbi'

distal_enhancers_ncbi=distal_enhancers
seqlevelsStyle(distal_enhancers_ncbi)='ncbi'

biomTrack = BiomartGeneRegionTrack(genome="hg38", biomart=ensembl,
                                   chromosome=thechr, 
                                   start=st, end=en,
                                   name="ENSEMBL",
                                   background.panel = "#FFFFFF",
                                   background.title = "#FFFFFF",
                                   col.axis="black",
                                   col.title="black",col=1,
                                   fill='gray70',
                                   ontfamily="Arial",
                                   showID = FALSE, geneSymbol = TRUE)


## ChIPseq
dtrac1 = DataTrack(range = bedgraph_dt_one_chr1,
                   type = "h",
                   genome = 'hg38',
                   name = "H3K4me3 HMD (40086)",
                   background.panel = "#FFFFFF",
                   background.title = "#FFFFFF",
                   col.axis="black",
                   col.title="black",col=1,
                   fill=1,
                   fontfamily="Arial" )

## peaks
thisReg_me3_peaks1 = AnnotationTrack(me3_peaks_ncbi, genome='hg38',
                                     name='H3K4me3',
                                     chromosome = thechr,shape='box',
                                     fill='blue',size=1.5,
                                     background.panel = "#FFFFFF",
                                     background.title = "#FFFFFF",
                                     col.axis="black",
                                     col.title="black",col=1,
                                     fontfamily="Arial")


thisReg_enhancers = AnnotationTrack(distal_enhancers_ncbi, genome='hg38',
                                    name='Distal Enhancers',
                                    chromosome = thechr,shape='box',
                                    fill='red',size=1.5, 
                                    background.panel = "#FFFFFF",
                                    background.title = "#FFFFFF",
                                    col.axis="black",
                                    col.title="black",col=1,
                                    fontfamily="Arial")


# itrack = IdeogramTrack(genome = "hg38", chromosome = thechr)
gtrack = GenomeAxisTrack()


plotTracks(list(dtrac1,
                thisReg_me3_peaks1, 
                thisReg_enhancers,
                biomTrack),
           chromosome = thechr, 
           from = st, to = en)

```


## H3K4me3 score at enhancers


Here, we will compute the H3K4me3 score from the bigWig file directly.


```{r}

compute_H3K4me3_score = function( bw_file, peak_file ){
  
  ## bw_file=me3_track; peak_file=distal_enhancers
  overlaps = findOverlaps( bw_file, peak_file )
  scores = elementMetadata( bw_file[ queryHits(overlaps) ])[,1]
  stopifnot(length(scores)==length(overlaps))
  os = unlist(lapply(split(scores,subjectHits(overlaps)), function(x){
    sum(x, na.rm=TRUE) }))
  res = data.frame( peak_number=as.numeric(names(os)),
                    peak_score=os,
                    stringsAsFactors=FALSE ) 
  return(res)}

```


Now integrate the H3K4me3 signal at all the defined enhancer classes. We will consider the HMD file for the best antibody


```{r, H3K4me3_HMD_CI_scores, eval=TRUE}

seqlevelsStyle(me3_track_HMD) = 'ucsc'
distal_enhancers_no_pol_me3_score = compute_H3K4me3_score( me3_track_HMD,
                                                           distal_enhancers_no_pol ) 

distal_enhancers_wi_pol_me3_score = compute_H3K4me3_score( me3_track_HMD,
                                                           distal_enhancers_with_pol ) 


distal_enhancers_no_exp_me3_score = compute_H3K4me3_score( me3_track_HMD,
                                                           distal_enhancers_no_expression_38 ) 

distal_enhancers_lo_exp_me3_score = compute_H3K4me3_score( me3_track_HMD,
                                                           distal_enhancers_low_expression_38 ) 

distal_enhancers_hi_exp_me3_score = compute_H3K4me3_score( me3_track_HMD,
                                                           distal_enhancers_high_expression_38 ) 


save( distal_enhancers_no_pol_me3_score, 
      distal_enhancers_wi_pol_me3_score, 
      distal_enhancers_no_exp_me3_score, 
      distal_enhancers_lo_exp_me3_score, 
      distal_enhancers_hi_exp_me3_score,
      file=paste0(objects, 'distal_enhancers_H3K4me3_scores_HMD.RData'))


```

Display the box-plots.

```{r, pol2_GRO-seq_distal_enhancer_H3K4me_score3, fig.width=12, fig.height=9, out.width='50%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

load( paste0( objects, 'distal_enhancers_H3K4me3_scores_HMD.RData') )


par(mfrow=c(1,2), mar=c(8,5,3,1), bty='n',lwd=2)
boxplot( log2(1+distal_enhancers_no_pol_me3_score[,2]), 
         log2(1+distal_enhancers_wi_pol_me3_score[,2]),
         outline=F, names=c('Absent','Present'), 
         col=c('red3','blue3'), xlab='Pol2',
         main='HMD file for Thermo', 
         ylab=expression("H3K4me3 (integrated signal (log"[2]*"))"), las=2)
boxplot( log2(1+distal_enhancers_no_exp_me3_score[,2]), 
         log2(1+distal_enhancers_lo_exp_me3_score[,2]),
         log2(1+distal_enhancers_hi_exp_me3_score[,2]),
         outline=F, 
         names=c('No','Low','High'), 
         col=c('red3','blue3','green3'),
         xlab='GRO-seq expression class', main='HMD file for Thermo',
         ylab=expression("H3K4me3 (signal (log"[2]*"))"), las=2)


t.test( log2(1+distal_enhancers_no_pol_me3_score[,2]),
        log2(1+distal_enhancers_wi_pol_me3_score[,2]) )

t.test( log2(1+distal_enhancers_no_exp_me3_score[,2]),
        log2(1+distal_enhancers_lo_exp_me3_score[,2]) )

t.test( log2(1+distal_enhancers_lo_exp_me3_score[,2]),
        log2(1+distal_enhancers_hi_exp_me3_score[,2]) )


```


Enhancers that overlap pol2 are more likely to overlap H3K4me3

```{r}

length( unique( queryHits(findOverlaps(distal_enhancers_no_pol, me3_peaks))) )
length(distal_enhancers_no_pol) - length( unique( queryHits(findOverlaps(distal_enhancers_no_pol, me3_peaks))) )
length( unique( queryHits(findOverlaps(distal_enhancers_with_pol, me3_peaks))) )
length(distal_enhancers_with_pol)-length( unique( queryHits(findOverlaps(distal_enhancers_with_pol, me3_peaks))) )

fisher.test( matrix( c(length( unique( queryHits(findOverlaps(distal_enhancers_no_pol, me3_peaks))) ),
                       length(distal_enhancers_no_pol)-length( unique( queryHits(findOverlaps(distal_enhancers_no_pol, me3_peaks))) ),
                       length( unique( queryHits(findOverlaps(distal_enhancers_with_pol, me3_peaks))) ),
                       length(distal_enhancers_with_pol)-length( unique( queryHits(findOverlaps(distal_enhancers_with_pol, me3_peaks))) )),
                     2, 2 ))

length( unique( queryHits(findOverlaps(distal_enhancers_no_pol, me3_peaks))) )/length(distal_enhancers_no_pol)

length( unique( queryHits(findOverlaps(distal_enhancers_with_pol, me3_peaks))) )/length(distal_enhancers_with_pol)

## expressed
eRNA_enhancers = c( distal_enhancers_low_expression_38, distal_enhancers_high_expression_38)

length( unique( queryHits(findOverlaps(distal_enhancers_no_expression_38, me3_peaks))) )/length(distal_enhancers_no_expression_38)

length( unique( queryHits(findOverlaps(eRNA_enhancers, me3_peaks))) )/length(eRNA_enhancers)


fisher.test( matrix( c(length( unique( queryHits(findOverlaps(distal_enhancers_no_expression_38, me3_peaks))) ),
                       length(distal_enhancers_no_expression_38)-length( unique( queryHits(findOverlaps(distal_enhancers_no_expression_38, me3_peaks))) ),
                       length( unique( queryHits(findOverlaps(eRNA_enhancers, me3_peaks))) ),
                       length(eRNA_enhancers)-length( unique( queryHits(findOverlaps(eRNA_enhancers, me3_peaks))) )),
                     2, 2 ))


```


## TBP and pol2-Ser5P


```{r}

tbp = readBed_filterChroms( paste0(peak_dir, "ENCFF609UTS.bed"), 
                                  chroms=paste0('chr',c(1:22,'X','Y')), 4 )

pol2ser5_rep1 = readBed_filterChroms(paste0(peak_dir,"GSM3536522_K562_PolSer5P_Rep1.bed"), 
                                  chroms=paste0('chr',c(1:22,'X','Y')), 3 )

pol2ser5_rep2 = readBed_filterChroms(paste0(peak_dir,"GSM3536522_K562_PolSer5P_Rep2.bed"), 
                                     chroms=paste0('chr',c(1:22,'X','Y')), 3 )


# pool the replicates
pol2ser5 = c( pol2ser5_rep1, pol2ser5_rep2 ) 


```



And assess the overlap between pol2 phosphorylated on Ser5 and with TBP.


```{r, warning=FALSE, Distal_enhancers_pol2Ser5_tbp, fig.width=10, fig.height=10, out.width='50%', fig.align='center', dev.args=list(pointsize=24),base.dir=figures, dev=c('png', 'pdf')}

distal_enhancers_hg19$pol2ser5 = countOverlaps( distal_enhancers_hg19, pol2ser5 )

hist( log2(distal_enhancers_hg19$pol2ser5),
      n=35, main='Pol2-Ser5 score at distal DHS',
      col='wheat4',
      xlab = expression('CUT&RUN (log2)') )
abline( v = 4, col='red4', lwd=3 )

export.bed( distal_enhancers_hg19[which(distal_enhancers_hg19$pol2ser5<4)],
            con = paste0(peak_dir, 'distal_enhancers_hg19_withoutPol2Ser5.bed' ) )

export.bed( distal_enhancers_hg19[which(distal_enhancers_hg19$pol2ser5>=4)],
            con = paste0(peak_dir, 'distal_enhancers_hg19_withPol2Ser5.bed' ) )


# ---------------------------------

distal_enhancers_GRCh38_pol2Ser5 = readBed_filterChroms(paste0(peak_dir,"distal_enhancers_GRCh38_withPol2Ser5.bed"), 
                                     chroms=paste0('chr',c(1:22,'X','Y')), 3 )

distal_enhancers_GRCh38_wo_pol2Ser5 = readBed_filterChroms(paste0(peak_dir,"distal_enhancers_GRCh38_withoutPol2Ser5.bed"), 
                                     chroms=paste0('chr',c(1:22,'X','Y')), 3 )

## define the classes of enhancers
distal_enhancers_GRCh38_wo_pol2Ser5_noTBP = distal_enhancers_GRCh38_wo_pol2Ser5[-unique(queryHits(findOverlaps(distal_enhancers_GRCh38_wo_pol2Ser5,tbp)))]

distal_enhancers_GRCh38_wo_pol2Ser5_withTBP = distal_enhancers_GRCh38_wo_pol2Ser5[unique(queryHits(findOverlaps(distal_enhancers_GRCh38_wo_pol2Ser5,tbp)))]

distal_enhancers_GRCh38_pol2Ser5_no_tbp = distal_enhancers_GRCh38_pol2Ser5[ -queryHits( findOverlaps(distal_enhancers_GRCh38_pol2Ser5,tbp)) ]

distal_enhancers_GRCh38_pol2Ser5_tbp = distal_enhancers_GRCh38_pol2Ser5[ queryHits( findOverlaps(distal_enhancers_GRCh38_pol2Ser5,tbp)) ]



statsP2 = c( length(unique(queryHits(findOverlaps(distal_enhancers_GRCh38_wo_pol2Ser5_noTBP,me3_peaks)))),
             length(unique(queryHits(findOverlaps(distal_enhancers_GRCh38_pol2Ser5_no_tbp,me3_peaks)))),
             length(unique(queryHits(findOverlaps(distal_enhancers_GRCh38_wo_pol2Ser5_withTBP,me3_peaks) ))),
             length(unique(queryHits(findOverlaps(distal_enhancers_GRCh38_pol2Ser5_tbp,me3_peaks) ))) )

statsP2

statsP2 = statsP2/c(length(distal_enhancers_GRCh38_wo_pol2Ser5_noTBP),
                    length(distal_enhancers_GRCh38_pol2Ser5_no_tbp),
                    length(distal_enhancers_GRCh38_wo_pol2Ser5_withTBP),
                    length(distal_enhancers_GRCh38_pol2Ser5_tbp))
statsP2 = statsP2*100

barplot( statsP2, col=c('wheat1','red3','orange3','gray60','blue'),
         names=c('Ser5P-TBP-','Ser5P+TBP-', 'Ser5P-TBP+', 'Ser5+TBP+' ), 
         ylim=c(0,40),
         ylab='% of distal enhancers intersecting H3K4me3 peak', 
         main = 'Signature of the DHS', las=2 )

save( distal_enhancers_GRCh38_wo_pol2Ser5_noTBP, 
      distal_enhancers_GRCh38_wo_pol2Ser5_withTBP, 
      distal_enhancers_GRCh38_pol2Ser5_no_tbp, 
      distal_enhancers_GRCh38_pol2Ser5_tbp,
      file=paste0(objects, 'Ser5_TBP_distal_enhancers_simplified.RData' ) )

```


Level of H3K4me3 at enhancers stratified based on Pol2-Ser5 phosphorylation and TBP binding.

```{r}

distal_enhancers_GRCh38_wo_pol2Ser5_noTBP_me3_score = 
  compute_H3K4me3_score( me3_track_HMD,
                         distal_enhancers_GRCh38_wo_pol2Ser5_noTBP ) 

distal_enhancers_GRCh38_wo_pol2Ser5_withTBP_me3_score = 
  compute_H3K4me3_score( me3_track_HMD,
                         distal_enhancers_GRCh38_wo_pol2Ser5_withTBP )

distal_enhancers_GRCh38_pol2Ser5_no_tbp_me3_score = 
  compute_H3K4me3_score( me3_track_HMD,
                         distal_enhancers_GRCh38_pol2Ser5_no_tbp )

distal_enhancers_GRCh38_pol2Ser5_tbp_me3_score = 
  compute_H3K4me3_score( me3_track_HMD,
                         distal_enhancers_GRCh38_pol2Ser5_tbp )


save( distal_enhancers_GRCh38_wo_pol2Ser5_noTBP_me3_score,
      distal_enhancers_GRCh38_wo_pol2Ser5_withTBP_me3_score, 
      distal_enhancers_GRCh38_pol2Ser5_no_tbp_me3_score,
      distal_enhancers_GRCh38_pol2Ser5_tbp_me3_score,
      file=paste0(objects,'H3K4me3_HMD_scores_pol2ser5_tbp_FINAL.RData' ) )

```

Display H3K4me3 level at enhancers striatifed by the presence of TBP and Pol2-Ser5P.


```{r, H3K4me3_score_enahncers_TBP_Pol2_ser5P, fig.width=9, fig.height=9, out.width='50%', fig.align='center', dev.args=list(pointsize=24),base.dir=figures, dev=c('png', 'pdf')}

load( paste0( objects, 'H3K4me3_HMD_scores_pol2ser5_tbp_FINAL.RData' ) )

par(mfrow=c(1,1), mar=c(8,5,3,1), bty='n',lwd=2)
boxplot( log2(0.1+distal_enhancers_GRCh38_wo_pol2Ser5_noTBP_me3_score[,2]),
         log2(0.1+distal_enhancers_GRCh38_pol2Ser5_no_tbp_me3_score[,2]), 
         log2(0.1+distal_enhancers_GRCh38_wo_pol2Ser5_withTBP_me3_score[,2]),
         log2(0.1+distal_enhancers_GRCh38_pol2Ser5_tbp_me3_score[,2]),
         outline=F, col=c('wheat3','red3','orange3', 'gray60'),
         names=c('Ser5P-TBP-','Ser5P+TBP-', 'Ser5P-TBP+','Ser5+TBP+' ),
         xlab='',
         main='', ylim=c(0,12),
         ylab=expression("H3K4me3 (integrated signal (log"[2]*"))"), las=2)

t.test( log2(0.1+distal_enhancers_GRCh38_wo_pol2Ser5_noTBP_me3_score[,2]),
         log2(0.1+distal_enhancers_GRCh38_pol2Ser5_no_tbp_me3_score[,2]) )

t.test( log2(0.1+distal_enhancers_GRCh38_pol2Ser5_no_tbp_me3_score[,2]),
         log2(0.1+distal_enhancers_GRCh38_wo_pol2Ser5_withTBP_me3_score[,2]) )


t.test( log2(0.1+distal_enhancers_GRCh38_wo_pol2Ser5_withTBP_me3_score[,2]),
         log2(0.1+distal_enhancers_GRCh38_pol2Ser5_tbp_me3_score[,2]) )

t.test( log2(0.1+distal_enhancers_GRCh38_wo_pol2Ser5_noTBP_me3_score[,2]),
         log2(0.1+distal_enhancers_GRCh38_pol2Ser5_tbp_me3_score[,2]) )

```


# GRO-cap based classification

Analysis of the GRO-cap data. This is Hg19.

```{r, warning=FALSE}

grocap_min = import.bw( paste0(wig_dir,"GSM1480321_K562_GROcap_wTAP_plus.bigWig") )
grocap_plus= import.bw( paste0(wig_dir,"GSM1480321_K562_GROcap_wTAP_minus.bigWig") )

grocap = c(grocap_min,grocap_plus)

```

Now score for expression level at the distal enhancers


```{r, warning=FALSE, distal_enhancers_GROCAP_signal, fig.width=15, fig.height=10, out.width='80%', fig.align='center', dev.args=list(pointsize=24),base.dir=figures, dev=c('png', 'pdf')}

## enhancers that do not overlap GRO-cap reads
distal_enhancers_no_grocap_expression = distal_enhancers_hg19[ -unique(queryHits(findOverlaps(distal_enhancers_hg19,grocap)))]

grocap_score_distal_enhancers = as.data.frame( findOverlaps(distal_enhancers_hg19,grocap) )
grocap_score_distal_enhancers$grocap = score(grocap)[ grocap_score_distal_enhancers[,2] ]
grocap_score_distal_enhancers = grocap_score_distal_enhancers[grocap_score_distal_enhancers$grocap>0,]
grocap_score_distal_enhancers = split(grocap_score_distal_enhancers[,3], grocap_score_distal_enhancers[,1])
distal_enhancers_hg19_grocap_transcription = unlist(lapply(grocap_score_distal_enhancers,sum))


hist( log2(distal_enhancers_hg19_transcription),n=50, 
      main='Enhancers with GRO-cap signal', col='gray60',
      xlab=expression("GRO-cap number of reads (log"[2]*")"))
abline(v=log2(5),col='red',lwd=2)

## enhancers with low transcription
distal_enhancers_low_grocap_expression = distal_enhancers_hg19[ as.numeric(names(distal_enhancers_hg19_grocap_transcription[distal_enhancers_hg19_grocap_transcription<=5]))]
## enhancers with high transcription
distal_enhancers_high_grocap_expression = distal_enhancers_hg19[ as.numeric(names(distal_enhancers_hg19_grocap_transcription[distal_enhancers_hg19_grocap_transcription>5]))]


```


Export the files and lift them over to GRCh38.


```{r, warning=FALSE}

export.bed(distal_enhancers_no_grocap_expression,
           paste0(peak_dir, "distal_enhancers_no_expression_GRO-cap_GRCh37_TAP_FINAL.bed") )
export.bed(distal_enhancers_low_grocap_expression,
           paste0(peak_dir, "distal_enhancers_low_expression_GRO-cap_GRCh37_TAP_FINAL.bed") )
export.bed(distal_enhancers_high_grocap_expression,
           paste0(peak_dir, "distal_enhancers_high_expression_GRO-cap_GRCh37_TAP_FINAL.bed") )

##################################################################

distal_enhancers_no_expression_38=readBed_filterChroms(paste0(peak_dir,"distal_enhancers_no_expression_GRO-cap_GRCh38_TAP_FINAL.bed"), chroms=paste0('chr',c(1:22,'X','Y')), 3)

distal_enhancers_low_expression_38=readBed_filterChroms(paste0(peak_dir,"distal_enhancers_low_expression_GRO-cap_GRCh38_TAP_FINAL.bed"), chroms=paste0('chr',c(1:22,'X','Y')), 3)

distal_enhancers_high_expression_38=readBed_filterChroms(paste0(peak_dir,"distal_enhancers_high_expression_GRO-cap_GRCh38_TAP_FINAL.bed"), chroms=paste0('chr',c(1:22,'X','Y')), 3)

save( distal_enhancers_high_expression_38,
      distal_enhancers_low_expression_38,
      distal_enhancers_no_expression_38,
      file= paste0(objects, 'Enhancers_GRO_stratification_GRO-cap_GRCh38_TAP.RData' ) )

```


How many of the distal DHS sites (stratified by expression measured by GRO-cap), overlap H3K4me3 peaks.


```{r, warning=FALSE, distal_enhancers_overlap_with_H3K4_GRO_CAP, fig.width=10, fig.height=8, out.width='50%', fig.align='center', dev.args=list(pointsize=24),base.dir=figures, dev=c('png', 'pdf')}

stats = c( length(unique(queryHits(findOverlaps(distal_enhancers_no_expression_38,me3_peaks)))),
           length(unique(queryHits(findOverlaps(distal_enhancers_low_expression_38,me3_peaks) ))),
           length( unique(queryHits(findOverlaps(distal_enhancers_high_expression_38,me3_peaks) )) ))

stats

stats = stats/c(length(distal_enhancers_no_expression_38),
                length(distal_enhancers_low_expression_38),
                length(distal_enhancers_high_expression_38))
stats = stats*100

barplot( stats, col=c('red','blue','blue4'),
         names=c('No','Low','High'), ylim=c(0,20),
         ylab='% of distal enhancers intersecting H3K4me3 peak', 
         xlab='GRO-cap expression level' )

```



```{r, H3K4me3_HMD_CI_scores_CAP, eval=TRUE}



distal_enhancers_no_exp_HMD_me3_score = compute_H3K4me3_score( me3_track_HMD,
                                                           distal_enhancers_no_expression_38 ) 

distal_enhancers_lo_exp_HMD_me3_score = compute_H3K4me3_score( me3_track_HMD,
                                                           distal_enhancers_low_expression_38 ) 

distal_enhancers_hi_exp_HMD_me3_score = compute_H3K4me3_score( me3_track_HMD,
                                                           distal_enhancers_high_expression_38 ) 


save( distal_enhancers_no_exp_HMD_me3_score, 
      distal_enhancers_lo_exp_HMD_me3_score, 
      distal_enhancers_hi_exp_HMD_me3_score,
      file=paste0(objects, 'distal_enhancers_H3K4me3_scores_HMD_GRO-Cap.RData'))


boxplot( log2(0.1+distal_enhancers_no_exp_HMD_me3_score[,2]), 
         log2(0.1+distal_enhancers_lo_exp_HMD_me3_score[,2]),
         log2(0.1+distal_enhancers_hi_exp_HMD_me3_score[,2]),
         outline=F, 
         names=c('No','Low','High'), 
         col=c('red3','blue3','green3'),
         main='GRO-cap (with TAP)',ylim=c(0,12),
         ylab=expression("H3K4me3 (integrated signal (log"[2]*"))"), las=2)

```



# Clustering of H3K4me2 signal around the TSS


We will consider the list of RefSeq genes imported above. We will focus on RefSeq identifiers and find promoter regions of genes (-2 to +8 kb). 

Define set of promoters to work with. Genes might have several RefSeq mRNA IDs, thus to decide which one to take, we will filter loci on the basis of gene symbol (like in our previous paper Pekowska *et al.*, *Genome Research* 2010). If there are several clustered TSS (within 200bp window), we will consider the one located at the shortest distance from the TTS. We keep only transcripts longer than 8kb.

The paper by Shah et al., was published in 2018, back when we first submitted our letter to the editor we used GRCh38.p12 version of the human genome assembly.

Find promoters that overlap only one gene. We identify promoters that do not overlap other genes and generate *GRanges* object with regions spanning -2 to +8 kb around the TSS. 


```{r}
genes = GRanges(seqnames = Rle(as.character(loci$chr)),
                 ranges = IRanges(as.numeric(loci$start_position),
                                  end = as.numeric(loci$end_position),
                                  names = seq(1, nrow(loci))),
                 strand = Rle(rep("*", nrow(loci))))
genes$gene = loci$refseq_mrna
genes$gene_name = loci$external_gene_name
genes$TSS=loci$transcription_start_site
genes$biotype=loci$gene_biotype
genes$STRAND=loci$STRAND


G_list = loci
G_list = G_list[G_list$gene_biotype=='protein_coding',]
G_list = G_list[G_list$refseq_mrna!="",]
G_list$RefSeq_curated = unlist(lapply( strsplit(G_list$refseq_mrna,"_"), function(x){x[[1]]} ))=="NM"
G_list = G_list[G_list$RefSeq_curated,]

gslm = split( G_list, G_list$external_gene_name ) 
table(unlist(lapply(gslm,function(x){length(unique(x$strand))}))) 
table(unlist(lapply(gslm, nrow))) 


## filter promoters
filtered_promoters = do.call('rbind', lapply( gslm, function(x){
  
  this_gene_strand = as.character( unique(x$STRAND) )
  not_too_wide = abs( max(x$transcription_start_site) - min( x$transcription_start_site ) ) < 200
  only_one_strand = length(this_gene_strand)==1
  protein_coding = unique( x$gene_biotype ) == 'protein_coding'
  
  tss = ifelse( this_gene_strand=='+', max(x$transcription_start_site), min(x$transcription_start_site) )
  tts = ifelse( this_gene_strand=='+', x$transcript_end[which.max(x$transcription_start_site)], x$transcript_start[which.min(x$transcription_start_site)] )
  chr = unique(x$chr)
  refseq = ifelse( this_gene_strand=='+', x$refseq_mrna[which.max(x$transcription_start_site)], x$refseq_mrna[which.min(x$transcription_start_site)] )
  
  START = ifelse( this_gene_strand=='+', tss-2000, tss-8000 )
  END = ifelse( this_gene_strand=='+', tss+2000, tss+2000 )
  long_enough = abs(tts-tss)>8000
  
  if( sum(not_too_wide,only_one_strand,protein_coding,long_enough) == 4 ) {
    
  data.frame( chr            = chr,
              prom_reg_start = START,
              prom_reg_end   = END,
              gene           = refseq, 
              gene_symbol    = unique( x$external_gene_name ), 
              tss            = tss,
              strand         = this_gene_strand,
              start_position = START,
              end_position   = END,
              stringsAsFactors = F) } } ) )
promoters = GRanges(seqnames = Rle(as.character(filtered_promoters$chr)),
                 ranges = IRanges(as.numeric(filtered_promoters$prom_reg_start),
                                  end = as.numeric(filtered_promoters$prom_reg_end),
                                  names = seq(1, nrow(filtered_promoters))),
                 strand = Rle(rep("*", nrow(filtered_promoters))))
names(promoters)=rownames(filtered_promoters)
promoters$gene = filtered_promoters$gene
promoters$gene_symbol = filtered_promoters$gene_symbol
promoters$STRAND = filtered_promoters$strand
promoters$TSS = filtered_promoters$tss

ovlp = findOverlaps(promoters,genes) 

mappings = data.frame( promoters = queryHits(ovlp),
                       gene_of_the_promoter_region = elementMetadata(promoters[queryHits(ovlp)])[,2],
                       overlapping_genes_geneName = elementMetadata(genes[subjectHits(ovlp)])[,2],
                       stringsAsFactors=FALSE )

length(unique(mappings$gene_of_the_promoter_region))
length(unique(mappings$promoters))

length(unique(mappings$overlapping_genes_geneName))

promoters_not_overlapping_other_genes = unlist( lapply( split( mappings, mappings$gene_of_the_promoter_region ),
                                                        function(x){
                                                          all(x$gene_of_the_promoter_region == x$overlapping_genes_geneName ) }))
promoters_not_overlapping_other_genes = promoters_not_overlapping_other_genes[which(promoters_not_overlapping_other_genes)]
table(promoters_not_overlapping_other_genes)

final_promoters = promoters[ which(names(promoters) %in% names(promoters_not_overlapping_other_genes)) ]


rss_rs = do.call('rbind', lapply( as.list(seq(1:length(final_promoters))), function(i){
  
  g = final_promoters[i]
  tss = g$TSS
  chr = as.character( chrom(g) )
  gene = names(g)
  if( g$STRAND=='+' ) { 
    return( 
      data.frame( chr = rep( (chr), 101),
                  starts = seq( tss-2000, tss+8000, length.out=101),
                  ends =  seq(tss-1901, tss+8099, length.out=101),
                  gene = rep( gene, 101) ) ) }
  else { 
  return( 
      data.frame( chr=rep( (chr), 101),
                  starts = rev( seq( tss-8099, tss+1901, length.out=101 ) ),
                  ends = rev( seq(tss-8000, tss+2000, length.out=101)),
                  gene = rep( gene, 101) ) ) }
} ) )



PsDF = GRanges(seqnames = Rle(rss_rs$chr),
                 ranges = IRanges(as.numeric(rss_rs$starts),
                                  end = as.numeric(rss_rs$ends),
                                  names = seq(1, nrow(rss_rs))),
                 strand = Rle(rep("*", nrow(rss_rs))))
PsDF$gene=(rss_rs$gene)
length(final_promoters) == length(PsDF)/101

save( final_promoters, PsDF, 
      file=paste0(objects,'final_promoters_and_ranges.RData') )



```


We will also map the gene symbols to ensembl gene IDs (we will need it in the GO-term enrichment analysis).


```{r, eval=FALSE}

filt=unique( PsDF$gene )

ensg = getBM(attributes= c("refseq_mrna",
                           "external_gene_name",
                           "ensembl_gene_id"),
             filters=c("external_gene_name"),
             values=filt, 
             mart=ensembl)
save( ensg, file=paste0(objects, 'ENSEMBL_genes_for_Average_profiles.Rdata') )

```


Now we can assess the signal of modifications. First define a function that will sum up the signal in bins defined above.


```{r}

getSignalInBins = function( bins, modification ){
  PsDFvsX = findOverlaps(bins,modification)
  PsDFvsX_score = split( modification$score[subjectHits(PsDFvsX)], queryHits(PsDFvsX) )
  PsDFvsX_score = unlist(lapply(PsDFvsX_score,function(x){sum(x,na.rm=T)}))
  
  bm = rep( 0, length(PsDF) )
  bm[as.numeric(names(PsDFvsX_score))]=PsDFvsX_score
  bm = matrix(bm,nrow=length(PsDF)/101, ncol=101, byrow = T )
  rownames(bm) = (PsDF$gene)[seq(1,length(PsDF),by=101)]
  
  return(bm)
}

```


### K562 - encode data replicates treated separetly

We consider encode data first, we analyze the two replicates separately.

```{r,eval=FALSE}
load(paste0(objects, 'ENSEMBL_genes_for_Average_profiles.Rdata'))
load( paste0(objects, 'final_promoters_and_ranges.RData') )
me2_encode1 = import( paste0(wig_dir,'ENCFF547SYT.bigWig') )
me2_encode2 = import( paste0(wig_dir,'ENCFF564BZY.bigWig') )

me2_encode1_pattern = getSignalInBins( PsDF, me2_encode1 )
write.table( me2_encode1_pattern, 
             file=paste0(objects, 'ENCFF547SYT_geneMatrix.txt'),
             sep='\t', quote=FALSE )

me2_encode2_pattern = getSignalInBins( PsDF, me2_encode2 )
write.table( me2_encode2_pattern, 
             file=paste0(objects, 'ENCFF564BZY_geneMatrix.txt'),
             sep='\t', quote=FALSE )

```


### K562 - ThermoFisher antibody (710796)

Here we process the bigwig profiles obtained with the antibody displaying the highest specificity according to Shah *et al.*, *Molecular Cell* 2018. We here consider data with no adjustments to show that they reproduce encode and they reproduce what the HMD showed in Shah et al., 2018.


```{r, eval=FALSE}

me2_best=import(paste0(wig_dir,'GSM2773400_AR16-5_TF-710796_mQ20_L200_genome_coverage.bigwig'))
me2_best_pattern = getSignalInBins( PsDF, me2_best )
write.table( me2_best_pattern, 
             file=paste0(objects, 'GSM2773400_AR16-5_TF-710796_mQ20_L200_genome_coverag_geneMatrix.txt'),
             sep='\t', quote=FALSE )


```


### K562 - the abcam antibody

Read the track files and perform the aggregation of signal. We also take the track from Shah et al., 2018 (GSE103543). 

Both ENCODE and Shah **use the same antibody - *ab7766* ** (ENCODE ID: ENCAB000ANF).

```{r, eval=FALSE}

me2_Shah=import(paste0(wig_dir,'GSM2773397_AR16-3_AB-7766_mQ20_L200_genome_coverage.bigwig'))
me2_Shah_pattern = getSignalInBins( PsDF, me2_Shah )
write.table( me2_Shah_pattern, 
             file=paste0(objects, 'GGSM2773397_AR16-3_AB-7766_mQ20_L200_genome_coverage_geneMatrix.txt'),
             sep='\t', quote=FALSE )


me2_encode_rep1_rep2 = import(paste0(wig_dir,'ENCFF491AUC.bigWig'))
me2_encode_pattern = getSignalInBins( PsDF, me2_encode_rep1_rep2 )
write.table( me2_encode_pattern, 
             file=paste0(objects, 'ENCFF491AUC_geneMatrix.txt'),
             sep='\t', quote=FALSE )

```


### K562 - the the normalized file from Shah et al. for the antibody used by the ENCODE consortium


```{r,eval=FALSE}
me2_HMD_ab7766=import(paste0( wig_dir, 'GSM2773397_AR16-3_AB-7766_mQ20_L200_HMD.bigwig') )
me2_HMD_ab7766_pattern = getSignalInBins( PsDF, me2_HMD_ab7766 )
write.table( me2_HMD_ab7766_pattern, 
             file=paste0(objects, 'GSM2773397_AR16-3_AB-7766_mQ20_L200_HMD_geneMatrix.txt'),
             sep='\t', quote=FALSE )
```

### K562 - the the normalized file from Shah et al.

Here we will consider the normalized file *GSE103543_AR16_H3K4me2-Corrected_HMD.bedgraph*. K562 Antibody: ab7766

```{r, eval=FALSE}
load(paste0(objects,'final_promoters_and_ranges.RData'))
me2_HMD=import(paste0( wig_dir, 'GSE103543_AR16_H3K4me2_Corrected_HMD.bigwig') )
me2_HMD_pattern = getSignalInBins( PsDF, me2_HMD )
write.table( me2_HMD_pattern, 
             file=paste0(objects, 'GSE103543_AR16_H3K4me2-Corrected_HMD_geneMatrix.txt'),
             sep='\t', quote=FALSE )
```



### B cells

Here the profiles were obtained using the ab7766 antibody.

```{r, eval=FALSE}

me2_Bcell_rep1_rep2 = import(paste0(wig_dir,'ENCFF945VIU.bigWig'))
me2_Bcell_pattern = getSignalInBins( PsDF, me2_Bcell_rep1_rep2 )
write.table( me2_Bcell_pattern, 
             file=paste0(objects, 'ENCFF945VIU_B_cell_rep1_rep2_geneMatrix.txt'),
             sep='\t', quote=FALSE )


```


## Clustering and GO-term analysis of H3K4me2 gene patterns 


Read the table and perform downstream clustering and GO-term analysis. We observed that in K562 a handful of genes displayed an unusually strong signal of H3K4me2. We will discard these genes from further analysis.

## Filtering

```{r, H3K4me2_pattern_data_filtering, fig.width=14, fig.height=12, out.width='70%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

load(paste0(objects, 'ENSEMBL_genes_for_Average_profiles.Rdata'))

me2_Shah_pattern = read.delim( file=paste0(objects, 'GGSM2773397_AR16-3_AB-7766_mQ20_L200_genome_coverage_geneMatrix.txt'),
                                   as.is=T, header=T )
me2_best_pattern = read.delim( file=paste0(objects, 'GSM2773400_AR16-5_TF-710796_mQ20_L200_genome_coverag_geneMatrix.txt'),
                                   as.is=T, header=T )
me2_HMD_pattern = read.delim( file=paste0(objects, 'GSE103543_AR16_H3K4me2-Corrected_HMD_geneMatrix.txt'),
                                   as.is=T, header=T )
me2_HMD_ab7766_pattern = read.delim( file=paste0(objects, 'GSM2773397_AR16-3_AB-7766_mQ20_L200_HMD_geneMatrix.txt'),
                                   as.is=T, header=T )

me2_encode_pattern = read.delim( file=paste0(objects, 'ENCFF491AUC_geneMatrix.txt'),
                                   as.is=T, header=T )

me2_encode_pattern1 = read.delim( file=paste0(objects, 'ENCFF547SYT_geneMatrix.txt'),
                                   as.is=T, header=T )
me2_encode_pattern2 = read.delim( file=paste0(objects, 'ENCFF564BZY_geneMatrix.txt'),
                                   as.is=T, header=T )



thr_me2_Shah_pattern = quantile( rowSums(me2_Shah_pattern), seq(0,1,by=0.001) )[991]
thr_me2_best_pattern = quantile( rowSums(me2_best_pattern), seq(0,1,by=0.001) )[991]
thr_me2_HMD_pattern = quantile( rowSums(me2_HMD_pattern), seq(0,1,by=0.001) )[991]
thr_me2_encode_pattern = quantile( rowSums(me2_encode_pattern), seq(0,1,by=0.001) )[991]
thr_me2_encode_pattern1 = quantile( rowSums(me2_encode_pattern1), seq(0,1,by=0.001) )[991]
thr_me2_encode_pattern2 = quantile( rowSums(me2_encode_pattern2), seq(0,1,by=0.001) )[991]

par( mfrow=c(3,2),mar=c(4,4,4,1), bty='n' )

plot(density(rowSums(me2_Shah_pattern)), main='Shah et al., abcam', 
     xlab='H3K4me2 integr. sig. (-2+8 kb TSS)', lwd=2, col='red4' )
abline(v=thr_me2_Shah_pattern, lty=2, col='gray')
plot(density(rowSums(me2_best_pattern)), main='Shah et al., Thermo', 
     xlab='H3K4me2 integr. sig. (-2+8 kb TSS)', lwd=2, col='red4' )
abline(v=thr_me2_best_pattern, lty=2, col='gray')
plot(density(rowSums(me2_HMD_pattern)), main='Shah et al., HMD', 
     xlab='H3K4me2 integr. sig. (-2+8 kb TSS)', lwd=2, col='red4' )
abline(v=thr_me2_HMD_pattern, lty=2, col='gray')
plot(density(rowSums(me2_encode_pattern)), main='Encode reps pooled', 
     xlab='H3K4me2 integr. sig. (-2+8 kb TSS)', lwd=2, col='red4' )
abline(v=thr_me2_encode_pattern, lty=2, col='gray')
plot(density(rowSums(me2_encode_pattern1)), main='Encode rep1', 
     xlab='H3K4me2 integr. sig. (-2+8 kb TSS)', lwd=2, col='red4' )
abline(v=thr_me2_encode_pattern1, lty=2, col='gray')
plot(density(rowSums(me2_encode_pattern2)), main='Encode rep2', 
     xlab='H3K4me2 integr. sig. (-2+8 kb TSS)', lwd=2, col='red4' )
abline(v=thr_me2_encode_pattern2, lty=2, col='gray')

```



Hereby, we identify genes for further clustering. We essentially removed 126 loci from further consideration.



```{r}

genes_2_remove = unique( c( rownames(me2_Shah_pattern)[rowSums(me2_Shah_pattern)>thr_me2_Shah_pattern],
                            rownames(me2_best_pattern)[rowSums(me2_best_pattern)>thr_me2_best_pattern],
                            rownames(me2_HMD_pattern)[rowSums(me2_HMD_pattern)>thr_me2_HMD_pattern],
                            rownames(me2_encode_pattern2)[rowSums(me2_encode_pattern2)>thr_me2_encode_pattern2] ) )

length(genes_2_remove)

```


Get the profiles for the relevant loci. 


```{r}

me2_Shah_pattern = me2_Shah_pattern[ ! rownames(me2_Shah_pattern) %in% genes_2_remove, ]
me2_best_pattern = me2_best_pattern[ ! rownames(me2_best_pattern) %in% genes_2_remove, ]
me2_HMD_pattern = me2_HMD_pattern[ ! rownames(me2_HMD_pattern) %in% genes_2_remove, ]
me2_HMD_ab7766_pattern = me2_HMD_ab7766_pattern[ ! rownames(me2_HMD_ab7766_pattern) %in% genes_2_remove, ]
me2_encode_pattern = me2_encode_pattern[ ! rownames(me2_encode_pattern) %in% genes_2_remove, ]
me2_encode_pattern1 = me2_encode_pattern1[ ! rownames(me2_encode_pattern1) %in% genes_2_remove, ]
me2_encode_pattern2 = me2_encode_pattern2[ ! rownames(me2_encode_pattern2) %in% genes_2_remove, ]

save( me2_Shah_pattern, me2_best_pattern, me2_HMD_pattern, 
      me2_encode_pattern, me2_encode_pattern1, me2_encode_pattern2,
      file = paste0(objects,'H3K4me2_profiles_filtereed.RData' )) 


```


### K562 - Shah data, abcam antibody

First the uncorrected values

```{r, eval=FALSE}
load( paste0(objects,'H3K4me2_profiles_filtereed.RData' ) )

clusters_Shah = kmeans(me2_Shah_pattern, 5, 100 )
genes = rownames(me2_Shah_pattern)
table(clusters_Shah$cluster)

save( clusters_Shah, genes,
      me2_Shah_pattern, file=paste0(objects, 'Shah_H3K4me2_clustering.RData') )

```


Plot the average profiles of the five clusters.


```{r, Shah_me2_clustering, fig.width=12, fig.height=10, out.width='50%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

load(paste0(objects, 'Shah_H3K4me2_clustering.RData'))
table(clusters_Shah$cluster)

cols = c('black','red','gray','blue','orange')
par(bty='n', mar=c(5,5,3,1))
plot( x=seq(-2000, 8000, by=100 ),
      colMeans( me2_Shah_pattern[ which(clusters_Shah$cluster == 1), ]),
      ty='l', col=cols[1], lwd=5, ylim=c(0,1500), main='H3K4me2 Shah ab7766',
      ylab='H3K4me2 signal', xlab='Distance from TSS (bp)')
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_Shah_pattern[ which(clusters_Shah$cluster == 2), ]),
      col=cols[2], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_Shah_pattern[ which(clusters_Shah$cluster == 3), ]),
      col=cols[3], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_Shah_pattern[ which(clusters_Shah$cluster == 4), ]),
      col=cols[4], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_Shah_pattern[ which(clusters_Shah$cluster == 5), ]),
      col=cols[5], lwd=5)

legend( x=4000,y=1500, paste("Cluster",1:5), bty='n',       
        col=cols, lwd=3 )

```



### K562 - Shah data - ThermoFisher antibody


```{r, eval=FALSE}

load( paste0(objects,'H3K4me2_profiles_filtereed.RData' ) )
clusters_best = kmeans(me2_best_pattern, 5, 100 )
genes = rownames(me2_best_pattern)
save( me2_best_pattern, genes,
      clusters_best, file=paste0(objects, 'Shah_Thermo_H3K4me2_clustering.RData') )

```


Plot the average profiles of the five clusters.


```{r, Shah_thermo, fig.width=12, fig.height=10, out.width='50%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

load(paste0(objects, 'Shah_Thermo_H3K4me2_clustering.RData'))
table(clusters_best$cluster)
cols = c('black','red','gray','blue','orange')

par(bty='n', mar=c(5,5,3,1))
plot( x=seq(-2000, 8000, by=100 ),
      colMeans( me2_best_pattern[ which(clusters_best$cluster == 1), ]),
      ty='l', col=cols[1], lwd=5, ylim=c(0,2000),
      main='H3K4me2 Shah ThermoFisher',
      ylab='H3K4me2 signal', xlab='Distance from TSS (bp)')
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_best_pattern[ which(clusters_best$cluster == 2), ]),
      col=cols[2], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_best_pattern[ which(clusters_best$cluster == 3), ]),
      col=cols[3], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_best_pattern[ which(clusters_best$cluster == 4), ]),
      col=cols[4], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_best_pattern[ which(clusters_best$cluster == 5), ]),
      col=cols[5], lwd=5)

legend( x=4000,y=2000, paste("Cluster",1:5), bty='n',       
        col=c('black','red','gray','blue','orange'), lwd=3 )

```


### K562 - Shah data - normalized final H3K4me2 file (HMD)

```{r, eval=FALSE}

clusters_HMD = kmeans(me2_HMD_pattern, 5, 50 )
genes = rownames(me2_HMD_pattern)
save( clusters_HMD, 
      genes, 
      me2_HMD_pattern,
      file=paste0(objects, 'H3K4me2_HMD_clustering.RData') )

```


Plot the average profiles of the five clusters.


```{r, Shah_normalized_HMD_all, fig.width=12, fig.height=10, out.width='50%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

load(paste0(objects, 'H3K4me2_HMD_clustering.RData'))
table(clusters_HMD$cluster)
cols = c('black','red','gray','blue','orange')

par(bty='n', mar=c(5,5,3,1))
plot( x=seq(-2000, 8000, by=100 ),
      colMeans( me2_HMD_pattern[ which(clusters_HMD$cluster == 1), ]),
      ty='l', col=cols[1], lwd=5, ylim=c(0,4000), main='H3K4me2 HMD file',
      ylab='H3K4me2 signal', xlab='Distance from TSS (bp)')
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_HMD_pattern[ which(clusters_HMD$cluster == 2), ]),
      col=cols[2], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_HMD_pattern[ which(clusters_HMD$cluster == 3), ]),
      col=cols[3], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_HMD_pattern[ which(clusters_HMD$cluster == 4), ]),
      col=cols[4], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_HMD_pattern[ which(clusters_HMD$cluster == 5), ]),
      col=cols[5], lwd=5)

legend( x=4000,y=4000, paste("Cluster",1:5), bty='n',       
        col=c('black','red','gray','blue','orange'), lwd=3 )


```


### K562 - Shah data - normalized  H3K4me2 file (HMD) for the best antibody

```{r, eval=FALSE}

clusters_HMD_ab7766 = kmeans(me2_HMD_ab7766_pattern, 5, 50 )
genes = rownames(me2_HMD_ab7766_pattern)
save( clusters_HMD_ab7766, 
      genes, 
      me2_HMD_ab7766_pattern,
      file=paste0(objects, 'H3K4me2_HMD_clustering_ab7766.RData') )

```


Plot the average profiles of the five clusters.


```{r, Shah_normalized_ab_7766, fig.width=12, fig.height=10, out.width='50%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

load(paste0(objects, 'H3K4me2_HMD_clustering_ab7766.RData'))
table(clusters_HMD$cluster)
cols = c('black','red','gray','blue','orange')

par(bty='n', mar=c(5,5,3,1))
plot( x=seq(-2000, 8000, by=100 ),
      colMeans( me2_HMD_ab7766_pattern[ which(clusters_HMD_ab7766$cluster == 1), ]),
      ty='l', col=cols[1], lwd=5, ylim=c(0,4000), main='H3K4me2 HMD ab7766',
      ylab='H3K4me2 signal', xlab='Distance from TSS (bp)')
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_HMD_ab7766_pattern[ which(clusters_HMD_ab7766$cluster == 2), ]),
      col=cols[2], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_HMD_ab7766_pattern[ which(clusters_HMD_ab7766$cluster == 3), ]),
      col=cols[3], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_HMD_ab7766_pattern[ which(clusters_HMD_ab7766$cluster == 4), ]),
      col=cols[4], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_HMD_ab7766_pattern[ which(clusters_HMD_ab7766$cluster == 5), ]),
      col=cols[5], lwd=5)

legend( x=4000,y=4000, paste("Cluster",1:5), bty='n',       
        col=c('black','red','gray','blue','orange'), lwd=3 )


```

### ENCODE data K562 

Both replicates

```{r, eval=FALSE}

clusters_encode = kmeans(me2_encode_pattern, 5, 50 )
genes = rownames(me2_encode_pattern)
save( clusters_encode, 
      genes, 
      me2_encode_pattern,
      file=paste0(objects, 'Encode_H3K4me2_clustering.RData') )

```

```{r, ENCODE_Rep1ll, fig.width=12, fig.height=10, out.width='50%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

load(paste0(objects, 'Encode_H3K4me2_clustering.RData'))
table(clusters_encode$cluster)
cols = c('black','red','gray','blue','orange')

par(bty='n', mar=c(5,5,3,1))
plot( x=seq(-2000, 8000, by=100 ),
      colMeans( me2_encode_pattern[ which(clusters_encode$cluster == 1), ]),
      ty='l', col=cols[1], lwd=5, ylim=c(0,2500), main='H3K4me2 ENCODE file (rep1 and rep2)',
      ylab='H3K4me2 signal', xlab='Distance from TSS (bp)')
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_encode_pattern[ which(clusters_encode$cluster == 2), ]),
      col=cols[2], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_encode_pattern[ which(clusters_encode$cluster == 3), ]),
      col=cols[3], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_encode_pattern[ which(clusters_encode$cluster == 4), ]),
      col=cols[4], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_encode_pattern[ which(clusters_encode$cluster == 5), ]),
      col=cols[5], lwd=5)

legend( x=4000,y=2500, paste("Cluster",1:5), bty='n',       
        col=c('black','red','gray','blue','orange'), lwd=3 )

```

Replicate 2.

```{r, eval=FALSE}

clusters_encode2 = kmeans(me2_encode_pattern2, 5, 50 )
genes = rownames(me2_encode_pattern2)
save( clusters_encode2, 
      genes, 
      me2_encode_pattern2,
      file=paste0(objects, 'Encode_rep2_H3K4me2_clustering.RData') )

```

```{r, ENCODE_Rep2, fig.width=12, fig.height=10, out.width='50%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

load(paste0(objects, 'Encode_rep2_H3K4me2_clustering.RData'))
table(clusters_encode2$cluster)
cols = c('black','red','gray','blue','orange')


par(bty='n', mar=c(5,5,3,1))
plot( x=seq(-2000, 8000, by=100 ),
      colMeans( me2_encode_pattern2[ which(clusters_encode2$cluster == 1), ]),
      ty='l', col=cols[1], lwd=5, ylim=c(0,1000), main='H3K4me2 ENCODE file Replicate 2',
      ylab='H3K4me2 signal', xlab='Distance from TSS (bp)')
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_encode_pattern2[ which(clusters_encode2$cluster == 2), ]),
      col=cols[2], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_encode_pattern2[ which(clusters_encode2$cluster == 3), ]),
      col=cols[3], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_encode_pattern2[ which(clusters_encode2$cluster == 4), ]),
      col=cols[4], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_encode_pattern2[ which(clusters_encode2$cluster == 5), ]),
      col=cols[5], lwd=5)


legend( x=4000,y=800, paste("Cluster",1:5), bty='n',       
        col=c('black','red','gray','blue','orange'), lwd=3 )


```


### B cell - encode data

```{r, eval=FALSE}
clusters_bcell = kmeans(me2_Bcell_pattern, 5, 100 )
genes = rownames(me2_Bcell_pattern)
save( clusters_bcell, 
      genes, 
      me2_Bcell_pattern,
      file=paste0(objects, 'Bcell_H3K4me2_clustering.RData') )

```

Plot the average profiles of the five clusters.

```{r, ENCODE_Bcells, fig.width=12, fig.height=10, out.width='50%', fig.align='center', dev.args=list(pointsize=24),fig.path=figures, dev=c('png', 'pdf')}

load(paste0(objects, 'Bcell_H3K4me2_clustering.RData'))
table(clusters_encode2$cluster)
cols = c('black','red','gray','blue','orange')


par(bty='n', mar=c(5,5,3,1))
plot( x=seq(-2000, 8000, by=100 ),
      colMeans( me2_Bcell_pattern[ which(clusters_bcell$cluster == 1), ]),
      ty='l', col=cols[1], lwd=5, ylim=c(0,2500), main='H3K4me2 Encode - B cell',
      ylab='H3K4me2 signal', xlab='Distance from TSS (bp)')
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_Bcell_pattern[ which(clusters_bcell$cluster == 2), ]),
      col=cols[2], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_Bcell_pattern[ which(clusters_bcell$cluster == 3), ]),
      col=cols[3], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_Bcell_pattern[ which(clusters_bcell$cluster == 4), ]),
      col=cols[4], lwd=5)
lines(x=seq(-2000, 8000, by=100 ),
      colMeans( me2_Bcell_pattern[ which(clusters_bcell$cluster == 5), ]),
      col=cols[5], lwd=5)

legend( x=4000,y=2500, paste("Cluster",1:5), bty='n',       
        col=c('black','red','gray','blue','orange'), lwd=3 )

```

## Analysis of chosen GO terms

We will compute P-values (Fisher) testing for the enrichment for the 4 main GO categories in the clustering of all the profiles.

```{r}

go_terms = c(multicellular_organismal_process='GO:0032501',
             developmental_process = 'GO:0032502',
             metabolic_process='GO:0008152',
             cellular_process='GO:0009987')


create_results_table = function( clustering_res, pattern, clusterID, ENSG, GO_terms ){

  ## clustering_res=clusters_Shah; pattern=me2_HMD_pattern;clusterID=1;GO_terms=go_terms;ENSG=ensg
  ## unique identifiers
  ENSG = ENSG[!duplicated(ENSG$external_gene_name),]
  ## universe and selection
  inUniverse = rownames(pattern)
  inUniverse = ensg[ensg$external_gene_name %in% inUniverse, 'ensembl_gene_id' ]
  inSelection =  names(clustering_res$cluster[ which(clustering_res$cluster == clusterID)] ) 
  inSelection = ensg[ensg$external_gene_name %in% inSelection, 'ensembl_gene_id' ]
  
  alg = factor( as.integer( inUniverse %in% inSelection ) )
  names(alg) = inUniverse
  
  tgd = new( "topGOdata",
           ontology="BP",
           allGenes = alg,
           nodeSize=10,
           annot=annFUN.org,
           mapping="org.Hs.eg.db",
           ID = "ensembl" )
  
  resultTopGO.classic = runTest(tgd, algorithm = "classic", statistic = "Fisher" )@score
  if( length(GO_terms)>0 ){
    return( resultTopGO.classic[names(resultTopGO.classic) %in% GO_terms ] )
  }
  if( length(GO_terms)==0 ){
    return( resultTopGO.classic )
  }
 
}

```



```{r, eval=TRUE}

load(paste0(objects, 'ENSEMBL_genes_for_Average_profiles.Rdata')) #
load(paste0(objects,'H3K4me2_profiles_filtereed.RData' )) #
load(paste0(objects, 'Shah_H3K4me2_clustering.RData')) 
load(paste0(objects, 'Shah_Thermo_H3K4me2_clustering.RData'))
load(paste0(objects, 'H3K4me2_HMD_clustering.RData')) #
load(paste0(objects, 'Encode_H3K4me2_clustering.RData')) #
load(paste0(objects, 'Encode_rep2_H3K4me2_clustering.RData')) #
load(paste0(objects, 'Bcell_H3K4me2_clustering.RData')) #


cluster_choice = c(me2_Shah_pattern=1, 
                   me2_best_pattern=5, 
                   me2_HMD_pattern=2, 
                   me2_HMD_ab7766_pattern=3, 
                   me2_encode_pattern=4, 
                   me2_encode_pattern2=3, 
                   me2_Bcell_pattern=1 )


shah_table_high = create_results_table( clustering_res = clusters_Shah,
                                        pattern = me2_Shah_pattern,
                                        clusterID = 1, ENSG = ensg, GO_terms = c() )  


shah_bestAb = create_results_table( clustering_res = clusters_best,
                                    pattern = me2_best_pattern,
                                    clusterID = 5,
                                    ENSG = ensg,
                                    GO_terms = c() )  


HMD_Shah = create_results_table( clustering_res = clusters_HMD,
                                 pattern = me2_HMD_pattern,
                                 clusterID = 2,
                                 ENSG = ensg,
                                 GO_terms = c() ) 

HMD_Shah_ab7766 = create_results_table( clustering_res = clusters_HMD_ab7766,
                                        pattern = me2_HMD_ab7766_pattern,
                                        clusterID = 3,
                                        ENSG = ensg,
                                        GO_terms = c() ) 

encode_table2 = create_results_table( clustering_res = clusters_encode2,
                                      pattern = me2_encode_pattern2,
                                      clusterID = 3,
                                      ENSG = ensg,
                                      GO_terms = c() )  


Bcell_table = create_results_table( clustering_res = clusters_bcell,
                                    pattern = me2_Bcell_pattern,
                                    clusterID = 1,
                                    ENSG = ensg,
                                    GO_terms = c() )  

## -----------------------------------------------

getGOnames = function( id_list, mappings ){
  # id_list=shah_table_high; mappings=goterms
  x=as.data.frame(id_list)
  colnames(x) = 'P-val'
  x$GO_name = mappings[ match(rownames(x), names(mappings))  ]
  x=x[order(x[,1], decreasing=FALSE),]
  return(x)
}


shah_table_high = getGOnames( shah_table_high, goterms )
shah_bestAb = getGOnames( shah_bestAb, goterms )
HMD_Shah = getGOnames( HMD_Shah, goterms )
HMD_ab776_Shah = getGOnames( HMD_Shah_ab7766, goterms )
encode_table2 = getGOnames( encode_table2, goterms )
Bcell_table = getGOnames( Bcell_table, goterms )

write.table( shah_table_high, file=paste0(objects, 'shah_table_high.txt'), quote=F, sep='\t' )
write.table( shah_bestAb, file=paste0(objects, 'shah_bestAb_high.txt'), quote=F, sep='\t' )
write.table( HMD_Shah, file=paste0(objects, 'HMD_Shah_high.txt'), quote=F, sep='\t' )
write.table( encode_table2, file=paste0(objects, 'encode_table2_high.txt'), quote=F, sep='\t' )
write.table( Bcell_table, file=paste0(objects, 'Bcell_table_high.txt'), quote=F, sep='\t' )


```


Using terms identified as within the top 4 terms in at least one sample, retrieve the P-values in all the other samples


```{r, FigureS1_panelB_GO, fig.width=8, fig.height=8, out.width='50%', fig.align='center', dev.args=list(pointsize=15),fig.path=figures, dev=c('png', 'pdf')}


load(paste0(objects, 'ENSEMBL_genes_for_Average_profiles.Rdata'))
load(paste0(objects,'H3K4me2_profiles_filtereed.RData' ))

shah_table_high = read.delim( paste0(objects, 'shah_table_high.txt'), as.is=TRUE )
shah_bestAb = read.delim( paste0(objects, 'shah_bestAb_high.txt'), as.is=TRUE )
HMD_Shah = read.delim( paste0(objects, 'HMD_Shah_high.txt'), as.is=TRUE )
encode_table2 = read.delim( paste0(objects, 'encode_table2_high.txt'), as.is=TRUE )
Bcell_table = read.delim( paste0(objects, 'Bcell_table_high.txt'), as.is=TRUE )

getTopGO = function(tb,topTerms){
  # tb=HMD_Shah; topTerms=4
  tbb = tb[1:topTerms,]
  res = rownames(tbb)
  names(res) = tbb[,'GO_name']
  return(res)}


cluster_choice

go_termsII = ( c( getTopGO(HMD_Shah,4),
                  getTopGO(shah_bestAb,4), 
                  getTopGO(encode_table2,4),
                  getTopGO(Bcell_table,4)) )

go_termsII = go_termsII[!duplicated(names(go_termsII))]


shah_bestAb = create_results_table( clustering_res = clusters_best,
                                    pattern = me2_best_pattern,
                                    clusterID = 5,
                                    ENSG = ensg,
                                    GO_terms = go_termsII )  


HMD_Shah = create_results_table( clustering_res = clusters_HMD,
                                 pattern = me2_HMD_pattern,
                                 clusterID = 2,
                                 ENSG = ensg,
                                 GO_terms = go_termsII ) 



encode_table2 = create_results_table( clustering_res = clusters_encode2,
                                      pattern = me2_encode_pattern2,
                                      clusterID = 3,
                                      ENSG = ensg,
                                      GO_terms = go_termsII )  


Bcell_table = create_results_table( clustering_res = clusters_bcell,
                                    pattern = me2_Bcell_pattern,
                                    clusterID = 1,
                                    ENSG = ensg,
                                    GO_terms = go_termsII )  




getGOnames = function( id_list, mappings ){
  # id_list=Bcell_table; mappings=goterms
  x=as.data.frame(id_list)
  colnames(x) = 'P-val'
  x$GO_name = mappings[ match(rownames(x), names(mappings))  ]
  return(x)
}


shah_bestAb1 = getGOnames( shah_bestAb, goterms )
HMD_Shah1 = getGOnames( HMD_Shah, goterms )
encode_table21 = getGOnames( encode_table2, goterms )
Bcell_table1 = getGOnames( Bcell_table, goterms )

orderGO = c("regulation of viral transcription","cellular protein metabolic process",
            "chromosome organization",
            "macromolecule catabolic process",
            "regulation of gene expression, epigenetic",
            "cellular macromolecule catabolic process",
            "hematopoietic or lymphoid organ development",
            "immune system development","immune system process")
orderGO = rev(orderGO)

resMatrix = cbind( Shah_Thermo = shah_bestAb1[ match(orderGO,shah_bestAb1$GO_name) ,1 ],
                   HMD_Shah1 = HMD_Shah1[ match(orderGO,HMD_Shah1$GO_name) ,1 ],
                   encode_table21 = encode_table21[ match(orderGO,encode_table21$GO_name) ,1 ],
                   Bcell_table1 = Bcell_table1[ match(orderGO,Bcell_table1$GO_name) ,1 ])
rownames(resMatrix) = orderGO

resMatrix_cut = matrix( cut(as.numeric(-log10(resMatrix)), 
                            c(1:14),
                            labels=F ),
                        nrow=nrow(resMatrix), 
                        ncol=ncol(resMatrix), 
                        byrow=FALSE )
rownames(resMatrix_cut) = rownames(resMatrix)
colnames(resMatrix_cut) = colnames(resMatrix)


par(mar=c(10,25,1,1),mfrow=c(1,1))
image( 1:4,1:nrow(resMatrix_cut),
       t(resMatrix_cut), axes=FALSE,xlab='',ylab='',
       col=colorRampPalette(c('white','red'))(12) )
box(col='black', lwd=2)
axis(1,at=1:4,c('Shah et al., Thermo',
                'Shah et al., HMD',
                'Encode K562', 'Encode B cell'), 
     las=2)
axis(2,at=1:nrow(resMatrix_cut),
     rownames(resMatrix_cut),las=1)
box(col='black')
```

And the color scale.

```{r,FigureS1_panelB_GO_colorScale, fig.width=8, fig.height=8, out.width='50%', fig.align='center', dev.args=list(pointsize=15),fig.path=figures, dev=c('png', 'pdf')}
par( mar=c(15,4,15,1)) 
image(0:length(c(1:14)), 0:1,
      matrix(c(1:14)), axes=FALSE,xlab='',ylab='',
      col=colorRampPalette(c('white','red'))(12) )
box(col='black', lwd=2)
axis(1,at=c(1:14) )
```
